Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-08-28 paper/ifo-complexAudio.tex
editing text in complex audio and wandering tone

We describe additional components to the demonstration in Section~\ref{sec:photodiode} and initial results in Section~\ref{sec:initialResultsOpMic}.
Filter and speech enhancement techniques applied to the data are detailed in Sections~\ref{sec:simple_filters} and~\ref{sec:logmmse} respectively.
This section is indirectly related to the advanced signal processing required to extract gravitational-wave signals. 
The optical microphone provides an interesting independent demonstration for a broader physics and engineering audience, particularly for the undergraduate lab.

\subsection{Photodiode}
\label{sec:photodiode}
% advanced method: explain how we capture data

Speech and music recordings require a higher sample rate than that of the webcam ($30\,{\rm Hz}$) used in Sections~\ref{sec:single_tone} and ~\ref{sec:viterbi_wandering}. 
Audible frequencies are in the range $\sim 20\,{\rm Hz}$--$20\,{\rm kHz}$. 
Speech intelligibility requires frequencies up to $3\,{\rm kHz}$ and music requires up to and beyond $8\,{\rm kHz}$~\cite{speech_intelligibility}. 
Therefore the optical microphone data requires a sample rate of at least $16\,{\rm kHz}$ to capture both speech and music. 
We achieved this with a photodiode.

A photodiode is an electrical component that acts as a regular diode when no light is incident on it, blocking any current flow in the `reverse' direction.
As the intensity of incident light rises, it becomes increasingly conductive in the reverse direction. 
% Fundamentally, a current is created in the photodiode by the photo-electric effect.
We place a OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp. 
This creates a photo-detector that produces a voltage dependent on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-centre position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}.
We see no difference when monitoring at other positions in the interference pattern.
The photodiode is mounted on a cloth screen re-purposed from the dismantled commercial speaker, with the electrical leads connected underneath.

The voltage signal from the photo-detector is captured by a MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi Model 3 v1.2~\cite{RaspberryPi:online}, which provides a convenient means to record the photodiode data.
Together, these sample the signal at roughly $16\,{\rm kHz}$. Resources for using the Pi are found in Appendix~\ref{app:circuit_diagram}.

@@

An updated schematic is shown in Figure~\ref{fig:ifo_schematic_podo} and a photograph of the entire optical microphone in Figure~\ref{fig:setup_pic2}.
See Appendix~\ref{app:circuit_diagram} for a diagram and photograph of the circuit.

@@

\subsection{Anti-aliased output}
\label{sec:initialResultsOpMic}
% breakdown of all the filters used along the way
The optical microphone was tested on recordings of varying speech from different people and on music ranging from simple melodies and rhythms to full songs. 
The source audio was played through the speaker with the optical microphone recording for the full duration of the audio (around a minute long for most sources). When processing the results, we often only looked at the first ten seconds of each observation for efficiency.
During recordings, care was taken to ensure minimal activity around the interferometer to reduce environmental noise. 
The timeseries was then directly converted to a .wav file and played as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}, as before.

We find that the raw output of the optical microphone, collected without any further processing (except anti-aliasing), is noisy, with a loud bass hum throughout.
Each $10\,{\rm s}$ Fourier spectrum shows dominant AC mains noise with power ranging from the fundamental of the $50\,{\rm Hz}$ mains power up to and beyond its $8$th harmonic, likely due to the non-linearity discussed in Appendix~\ref{app:intensity_derivation} (see also Ref.~\cite{feynman}). 
This mains power signal is present in the power spectral density of background noise as shown in the top panel of Figure~\ref{fig:psd_noise}.
The mains power signal is far weaker but still present in the background noise taken with the photodiode in darkness, suggesting that ambient lighting has a large contribution.
The origin of the mains power signal is likely to be from the photodiode signal, or environmental noise from the rooms lighting and air conditioning, or a combination. 
Outside the mains noise, the power appears to peak broadly around 750 Hz, the origin or significance of which we don't know.

@@

Fig.~\ref{fig:noiseSubtract} shows the frequency spectrum of $1\,{\rm s}$ of recorded speech. 
The first and second panels show the original signal before passing through the interferometer and the optical microphone recording of the signal respectively. 
An intuitive means to remove noise is to subtract the noise spectrum of the instrument directly from the signal spectrum. 
The third panel of Fig.~\ref{fig:noiseSubtract} shows a recording from the optical microphone with no audio signal. 
We see that the output noise is coloured, which reflects the non-linearity of the optical microphone.
The fourth panel of Fig.~\ref{fig:noiseSubtract} shows the frequency spectrum after the noise spectrum has been subtracted. 
We see no obvious improvement, which may be due to time-variant noise.

@@

We also apply a Butterworth filter. 
A low-pass Butterworth filter with cut-off frequency $f_c$ and gain $\varepsilon$, has a frequency response (attenuation at each frequency)
\begin{equation}
\label{eq:butterworth}
H(\omega) = \left(1+\varepsilon^2 \left( \frac{f}{f_c} \right)^{2n}\right)^{-\frac{1}{2}}
\end{equation}
% commenting to send to co-authors
%\jam{[Andrew wanted a quantitative comment about what the Butterworth does to the PSD, what do I use?]}
To construct a Butterworth band-pass filter, the low-pass filter in Eqn.~\ref{eq:butterworth} is combined with a high-pass filter of similar form. 
The best performing band-pass filter we find is a fifth order ($n = 5$) Butterworth filter over the range $150\,{\rm Hz}$--$3\,{\rm kHz}$ (beyond $2\,{\rm kHz}$ speech information content drops significantly~\cite{speech_intelligibility}). 
The effect of the Butterworth band-pass filter on the power spectral density of the background noise is shown in the bottom panel of Fig.~\ref{fig:psd_noise}, where we see that it reduces the amplitude of the mains harmonics below $150\,{\rm Hz}$ 
The response of the filter is shown in Fig.~\ref{fig:butterworth}.
Band-passing the signal does not address the remaining harmonics above $100\,{\rm Hz}$. 
However, it does suppress other noise at higher frequencies. 
The Butterworth filter performs well at maintaining source intelligibility as it is optimised to be `maximally flat' within the band region (prioritising condition ii) above).
However recordings of speech still have low intelligibility (the speech can't be understood).
Looking forward to Sec.~\ref{sec:logmmse}, results for the Butterworth filter are shown in combination with a speech enhancement technique to address this.

@@

\subsubsection{Cascade notched filter}
A straightforward method for removing hum noise and its harmonics is to use a sequence of notch filters centered at each of the frequency bins we want to remove. 
A typical transfer function of infinite impulse response (IIR) notch filter can be written as \citep{10.5555/541204}
\begin{equation}
    \label{eqn:notch}
    H(z)=\frac{1+\alpha}{2}\frac{1-2\beta z^{-1}+z^{-2}}{1-\beta(1+\alpha)z^{-1}+\alpha z^{-2}}
\end{equation}
where $\alpha$ and $\beta$ are self-tuned parameters which adjust the locations of poles and zeros. 
By allowing the numerator and demoninator of $H(z)$ to be zero, we can find the corresponding locations of zeros and poles of the filter. 
Notice zeros are at $e^{\pm jw_0}, \text{where } w_0=\cos^{-1}(\beta)$, and poles are at $re^{\pm j\phi}, \text{where } r=\sqrt{\alpha},\phi=\cos^{-1}(\beta(1+\alpha)/(2\sqrt{\alpha}))$.
At $w_0$, the magnitude response will be zero. 
Thus, we can calculate $\beta$ setting $w_0$ to be the frequency we want to remove. 
The parameter $\alpha$ affects the bandwidth ($B_w$) of the filter and has the relation $B_w=\cos^{-1}(\frac{2\alpha}{1+\alpha^2})$. \\
In our paper, we first decide $\beta$ for each filter. 
We locate hum noise spectrum at around $50k\,{\rm Hz}$ where the integer $k=1,2,\dots,15$. 
For each $k$, we build a notch filter to filter out the specific frequency bin approximately at $50k \, {\rm Hz}$. 
We then alter the parameter $\alpha$ to determine the bandwidth of each filter. 
On one side, the bandwidth should be as narrow as possible to avoid disturbing useful signals. 
This can be achieved through increasing orders of the filter (add more poles and zeros). 
On the other side, the bandwidth should have enough residues to account for the uncertainty of our pinpointing the location of hum noise. 
After considering these two aspects, we are now able to get the ideal transfer function for each filter. 
The whole cascaded filter is achieved by multiplying them together, 
\begin{equation}
    \label{eqn:notch15}
    H(z)=H_1(z)H_2(z)\dots H_k(z),k=1,2,\dots 15\,,
\end{equation}
where $H(z)$ denotes the transfer function of the full cascaded notch filter, and $H_k(z)$ denotes one single notch filter for removing $50k \, {\rm Hz}$ frequency. 
We use the built-in MATLAB filter design toolbox to help us realize a well performed cascaded notch filter (each filter has order 6).  
The magnitude response of the first five notch filters are shown in Fig.~\ref{fig:notchMagResponse}. 
\begin{figure}
\begin{center}
\includegraphics[width=.49\textwidth]{figures/1111}
\end{center}
\caption{\label{fig:notchMagResponse}
Magnitude response for the first five notches of the $15$ cascaded notch filter described by Eqns.~\ref{eqn:notch} and ~\ref{eqn:notch15}. 
}
\end{figure}

The timeseries and frequency spectrum results of the notch filter are shown in the third panel of Fig.~\ref{fig:timeNotchWiener} and the first panel of Fig.~\ref{fig:freqNotchWiener}. 
We seen that the mains harmonics are significantly attenuated in comparison to the analysis in Fig.~\ref{fig:noiseSubtract}.

@@

A Wiener filter is used to compute an estimate of the original signal. 

@@

Given the observation data $\textbf{x}$, our goal is to filter the signal $\textbf{s}$ from the noise $\textbf{w}$, and get an estimated $\hat{\textbf{s}}$, where the difference between $\hat{\textbf{s}}$ and true signal $\textbf{s}$ is minimized in the mean square error sense, which is represented as

@@

Assuming that (i) $\textbf{x}$ is wide sense stationary~\footnote{

@@

with zero mean; (ii) the signal $\textbf{s}$ is zero mean; (iii) the noise processes $\textbf{w}$ are uncorrelated with the signal $\textbf{s}$, we can further express the estimated $\hat{\textbf{s}}$ to be a linear combination of present and past input observed data

>>>>

We describe additional hardware components to the demonstration in Section~\ref{sec:photodiode} and initial results in Section~\ref{sec:initialResultsOpMic}.
Filter and speech enhancement techniques applied to the data are detailed in Sections~\ref{sec:simple_filters} and~\ref{sec:logmmse} respectively.
This section is indirectly related to the advanced signal processing required to extract gravitational-wave signals. 
The optical microphone provides an interesting independent demonstration for a broader physics and engineering audience, particularly for the undergraduate lab.

\subsection{Hardware modifications for the optical microphone}
\label{sec:photodiode}
% advanced method: explain how we capture data

Speech and music recordings require a higher sample rate than that of the webcam ($30\,{\rm Hz}$) used in Sections~\ref{sec:single_tone} and ~\ref{sec:viterbi_wandering}. 
Audible frequencies are in the range $\sim 20\,{\rm Hz}$--$20\,{\rm kHz}$. 
Speech intelligibility requires frequencies up to $3\,{\rm kHz}$ and music requires up to and beyond $8\,{\rm kHz}$~\cite{speech_intelligibility}. 
Therefore the optical microphone data requires a sample rate of at least $16\,{\rm kHz}$ to capture both speech and music. 
We achieved this with a photodiode\footnote{
A photodiode is an electrical component that acts as a regular diode when no light is incident on it, blocking any current flow in the `reverse' direction.
As the intensity of incident light rises, it becomes increasingly conductive in the reverse direction.
}

We place a OSRAM BPW21 photodiode in reverse-bias over an LM358 op-amp which makes a photo-detector that produces a voltage dependent on the incident intensity. 
The photodiode records the interference pattern at roughly the same off-centre position as the webcam in Sections~\ref{sec:single_tone} and~\ref{sec:viterbi_wandering}.
We see no difference when monitoring at other positions in the interference pattern.
The photodiode is mounted on a cloth screen re-purposed from the dismantled commercial speaker, with the electrical leads connected underneath. 
The voltage signal from the photo-detector is captured by a MCP3008 $10$-bit analog-to-digital converter (ADC) connected to a Raspberry Pi (Pi) Model 3 v1.2~\cite{RaspberryPi:online}, which provides a convenient means to record the photodiode data.
Together, these sample the signal at roughly $16\,{\rm kHz}$. 
Resources for using the Pi and photdiode are detailed in Appendix~\ref{app:circuit_diagram}.

@@

An updated schematic is shown in Fig.~\ref{fig:ifo_schematic_podo} and a photograph of the entire optical microphone in Fig.~\ref{fig:setup_pic2}.
See Appendix~\ref{app:circuit_diagram} for a diagram and photograph of the circuit.

@@

\subsection{Anti-aliased output}
\label{sec:initialResultsOpMic}

We test the optical microphone with a variety of recordings including speech from different people and music ranging from simple melodies and rhythms to full songs. 
%The source audio was played through the speaker with the optical microphone recording for the full duration of the audio (around a minute long for most sources). 
During recordings, care was taken to ensure minimal activity around the interferometer to reduce environmental noise. 
The timeseries was then directly converted to a .wav file and played as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}, as before.
When processing the results, we generally only analysed the first $10\,{\rm s}$ of each observation for efficiency.

The raw output of the optical microphone (with only anti-aliasing) is noisy with a loud bass hum throughout. 
Each $10\,{\rm s}$ Fourier spectrum shows dominant AC mains noise with power from the fundamental of the $50\,{\rm Hz}$ mains power up to and beyond its $8$th harmonic, likely due to the non-linearity discussed in Appendix~\ref{app:intensity_derivation} (see also Ref.~\cite{feynman}).
This mains power signal is present in the power spectral density of background noise, with the speaker switched off, as shown in the top panel of Fig.~\ref{fig:psd_noise}.
The mains signal is still present (but far weaker) in measurements with the photodiode in darkness, suggesting that ambient lighting has a large contribution.
Besides lighting, other possible contributions to the mains signal include air conditioning and the photodiode. 
There is also a broad peak visible at around $750\,{\rm Hz}$ in Fig.~\ref{fig:psd_noise}, which we have not been able to identify an origin for.

@@

Fig.~\ref{fig:noiseSubtract} shows the frequency spectrum of $1\,{\rm s}$ of recorded speech. 
The first and second panels show the original signal before passing through the interferometer and the optical microphone recording of the signal respectively. 
An intuitive means to remove noise is to subtract the noise spectrum of the instrument directly from the signal spectrum. 
The third panel of Fig.~\ref{fig:noiseSubtract} shows a recording from the optical microphone with no input audio signal. 
We see that the output noise is coloured, which reflects the non-linearity of the optical microphone.
The fourth panel of Fig.~\ref{fig:noiseSubtract} shows the frequency spectrum after the noise spectrum has been subtracted. 
We see no obvious improvement, which may be due to time-variant noise.

@@

We next apply a Butterworth filter. 
A low-pass Butterworth filter with cut-off frequency $f_c$ and gain $\varepsilon$, has a frequency response (attenuation at each frequency)
\begin{equation}
\label{eq:butterworth}
H(\omega) = \left[1+\varepsilon^2 \left( \frac{f}{f_c} \right)^{2n}\right]^{-1/2}
\end{equation}
% commenting to send to co-authors
%\jam{[Andrew wanted a quantitative comment about what the Butterworth does to the PSD, what do I use?]}
To construct a Butterworth band-pass filter, the low-pass filter in Eqn.~\ref{eq:butterworth} is combined with a high-pass filter of similar form. 
The best performing band-pass filter we find is a fifth order ($n = 5$) Butterworth filter over the range $150\,{\rm Hz}$--$3\,{\rm kHz}$ (beyond $2\,{\rm kHz}$ speech information content drops significantly~\cite{speech_intelligibility}). 
The effect of the Butterworth band-pass filter on the power spectral density of the background noise is shown in the bottom panel of Fig.~\ref{fig:psd_noise}, where we see that it reduces the amplitude of the mains harmonics below $150\,{\rm Hz}$ 
The response of the filter is shown in Fig.~\ref{fig:butterworth}.
Band-passing the signal does not address the remaining harmonics above $100\,{\rm Hz}$. 
However, it does suppress other noise at higher frequencies. 
In general, the Butterworth filter performs well at maintaining source intelligibility as it is optimised to be `maximally flat' within the band region (prioritising condition ii) above).
However recordings of speech still have low intelligibility (the speech can't be understood).
Looking forward to Sec.~\ref{sec:logmmse}, results for the Butterworth filter are shown in combination with a speech enhancement technique to address this.

@@

\subsubsection{Cascade notched filter}
A straightforward method for removing hum noise and its harmonics is to use a sequence of notch filters centered at each of the frequency bins we want to remove. 
A typical transfer function of infinite impulse response (IIR) notch filter can be written as \citep{10.5555/541204}
\begin{equation}
    \label{eqn:notch}
    H(z)=\frac{1+\alpha}{2}\frac{1-2\beta z^{-1}+z^{-2}}{1-\beta(1+\alpha)z^{-1}+\alpha z^{-2}}
\end{equation}
where $\alpha$ and $\beta$ are self-tuned parameters which adjust the locations of poles and zeros and $z$ is the complex frequency. 
By allowing the numerator and demoninator of $H(z)$ to be zero, we can find the corresponding locations of zeros and poles of the filter. 
There are zeros are at $e^{\pm jw_0}, \text{where } w_0=\cos^{-1}(\beta)$, and poles are at $re^{\pm j\phi}, \text{where } r=\sqrt{\alpha} \text{ and }\phi=\cos^{-1}[\beta(1+\alpha)/2\sqrt{\alpha}]$.
At $w_0$, the magnitude response will be zero. 
Thus, we can calculate $\beta$ setting $w_0$ to be the frequency we want to remove. 
The parameter $\alpha$ affects the bandwidth $B_w$ of the filter where $B_w=\cos^{-1}[2\alpha/(1+\alpha^2)]$. 
%In our paper, we first decide $\beta$ for each filter.

In this work we find a cascade filter with $15$ notches works well. 
We decide $\beta$ first by building a notch filter at each of the frequency bins located at $50k\,{\rm Hz}$ where the integer has values $k=1,2,\dots,15$. 
%The mains noise is located at $50k\,{\rm Hz}$ where the integer has values $k=1,2,\dots,15$. 
%For each $k$, we build a notch filter at the specific frequency bin at $50k \, {\rm Hz}$. 
We then alter the parameter $\alpha$ to determine the bandwidth of each filter. 
On one side, the bandwidth should be as narrow as possible to avoid disturbing useful signals. 
This can be achieved through increasing orders of the filter (add more poles and zeros). 
On the other side, the bandwidth should have enough residues to account for the uncertainty of our pinpointing the location of hum noise. 
After considering these two aspects, we find the ideal transfer function for each filter. 
The whole cascaded filter is achieved by multiplying them together, 
\begin{equation}
    \label{eqn:notch15}
    H(z) = H_{k=1}(z) H_{k=2}(z)\dots H_{k=15}(z)
%    H(z)=H_1(z)H_2(z)\dots H_k(z),k=1,2,\dots 15\,,
\end{equation}
where $H(z)$ denotes the transfer function of the full cascaded notch filter, and $H_k(z)$ denotes one single notch filter for removing $50k \, {\rm Hz}$ frequency. 
We use the built-in MATLAB filter design toolbox to help us realize a well performed cascaded notch filter (each filter has order 6).  
The magnitude response of the first five notch filters are shown in Fig.~\ref{fig:notchMagResponse}. 
\begin{figure*}
\begin{center}
\includegraphics[width=.7\textwidth]{figures/1111}
\end{center}
\caption{\label{fig:notchMagResponse}
Magnitude response for the first five notches of the $15$ cascaded notch filter described by Eqns.~\ref{eqn:notch} and ~\ref{eqn:notch15}. 
}
\end{figure*}

The timeseries and frequency spectrum results of the notch filter are shown in the third panel of Fig.~\ref{fig:timeNotchWiener} and the first panel of Fig.~\ref{fig:freqNotchWiener} respectively. 
We see that the mains harmonics are significantly attenuated in comparison to the analysis in Fig.~\ref{fig:noiseSubtract}.

@@

Here we use a Wiener filter to compute an estimate of the original signal. 

@@

Given the observation data $\textbf{x}$, our goal is to filter the signal $\textbf{s}$ from the noise $\textbf{w}$ to find an estimated $\hat{\textbf{s}}$, where the difference between $\hat{\textbf{s}}$ and true signal $\textbf{s}$ is minimized in the mean square error sense, which is represented as

@@

Assuming that: i) $\textbf{x}$ is wide sense stationary~\footnote{

@@

with zero mean; ii) the signal $\textbf{s}$ is zero mean; and iii) the noise processes $\textbf{w}$ are uncorrelated with the signal $\textbf{s}$, we can further express the estimated $\hat{\textbf{s}}$ to be a linear combination of present and past input observed data

### 2020-08-28 paper/ifo-wanderingTone.tex
editing text in complex audio and wandering tone

Continuous gravitational wave signals may wander slowly in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or stochastic accretion flows in binary neutron stars~\cite{BildstenTB:1998}. 
%As also mentioned in Section~\ref{sec:introduction}, continuous wave searches often look for signals that wander slowly in frequency~\cite{ScoX1O2Viterbi:2019}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch. 
%Unlike the single tone described in Section~\ref{sec:single_tone}, finding the signal isn’t as simple as searching for a peak in the spectrum for the full timeseries data.%observing run.

@@

In continuous wave searches a hidden Markov model is used to describe the system. 
The frequency (the state) of the signal is unknown (hidden) and the signal is Markovian in that its current state depends only on the previous state. 
The hidden Markov model method relates the observable quantity (the detection statistic) to the signal frequency (the hidden state). 
The task is then to find the most likely wandering frequency signal given the observable of the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
This method is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

\item The Viterbi path is then found by backtracking along the stored best paths at each time step (see also Appendix~\ref{app:viterbi}). 
% come back to this. 
%\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

To test the Viterbi algorithm, a sinusoidal signal with wandering frequency was played by the speaker and the output recorded via webcam as in Section~\ref{sec:single_tone}. The injected frequency itself followed a sinusoid with decaying amplitude. This choice was arbitrary, any injected path would have worked, as long as it changed slowly enough for the search parameters of the algorithm.
%Note that when creating a wandering frequency signal, the phase change must be integrated up over time, as detailed in Appendix~\ref{app:phase_gotcha}.

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively maximized over all terminating frequency bins. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
We also consider the root mean square (RMS) fractional error $E_{\rm rms}$ along the path. Given $i=0,1,...,N_f$ columns of the injected frequency $I_i$ and recovered (Viterbi) frequency $R_i$, $E_{\rm rms}$ is given by
\begin{equation}
E_{\rm rms} = \sqrt{ \frac{1}{N_f+1} \sum_{i=0}^{N_f} \frac{(I_i - R_i)^2}{{I_i}^2} }
\end{equation}
Computing this along the Viterbi path we find $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

%The Viterbi path stays within one frequency bin (approximately $0.3\,{\rm Hz}$) of the injected tone for $94\%$ of the run. 
Initially the injected signal moves faster then the frequency wander limit, leading to the discrepancy observed for $t\lesssim 10\,{\rm s}$. 
%Note that the signal initially moves faster than the frequency wander limit, causing a discrepancy between the injected and recovered signals. 
%To fix this, the limit should be doubled, although this would then cause the recovered path to stray more often. 
There is also an anomaly at $150\,{\rm s}$, which illustrates what happens when the interferometer is disturbed, e.g. when someone walks nearby. 
The Viterbi algorithm is robust against such disturbances, recovering the optimal path promptly, as Fig.~\ref{fig:viterbi_overlay} shows. 
%is likely due to a disturbance around the interferometer (such as someone walking nearby).
%There also appears to be an anomaly at $150\,{\rm s}$, likely from a disturbance (such as someone walking) around the interferometer.

%Empirically, decreasing the signal-to-noise ratio closer to unity (by making the speaker quieter) causes the Viterbi algorithm to stray more from the signal. 
%However, since the only correlation through the grid is the signal, we always expect the Viterbi path to lie near to the signal no matter how high the noise is, and this is indeed what we see. \han{Should we show results for this?}

>>>>

Continuous gravitational wave signals may wander slowly in frequency over time, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or stochastic accretion flows in binary neutron stars~\cite{BildstenTB:1998}. 
The audio analogue of this is a tone that wanders in frequency; a note that changes pitch.

@@

In continuous wave searches a hidden Markov model is used to describe the system. 
The frequency (the state) of the signal is unknown (hidden) and the signal is Markovian in that the state at any time depends only on the state in had at the previous time step. 
The hidden Markov model method relates the observable quantity (the detection statistic) to the signal frequency (the hidden state). 
The task is to find the most likely wandering frequency signal given the observable of the spectrogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
This method is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

\item The Viterbi path is then found by backtracking along the stored best paths at each time step (see also Appendix~\ref{app:viterbi}). 
% come back to this. 
%\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in Fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

To test the Viterbi algorithm, a sinusoidal signal with wandering frequency was played through the speaker and the output recorded via webcam as in Section~\ref{sec:single_tone}. 
The injected frequency itself followed a sinusoid with decaying amplitude. 
This choice was arbitrary, any injected path would have worked, as long as it changed slowly enough for the search parameters of the algorithm.

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot markers show the injected signal. 
The white-cross markers show the recovered Viterbi path maximised over all terminating frequency bins, which qualitatively provides a good match to the injection. 
Initially, the injected signal moves faster than the frequency wander limit, leading to the discrepancy observed for $t\lesssim 10\,{\rm s}$. 
There is also an anomaly at $150\,{\rm s}$, which illustrates what happens when the interferometer is disturbed, e.g. when someone walks nearby. 
The Viterbi algorithm is robust against such disturbances, recovering the optimal path promptly, as Fig.~\ref{fig:viterbi_overlay} shows.

Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
Another figure of merit is the root mean square (RMS) fractional error $E_{\rm rms}$ along the path defined as 
\han{}
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. 
Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay} we find $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.
